---
title: "Final Annotated R Output"
output: html_notebook
author: "Cole Maher"
date: ""
---

# Load packages

```{r}
pacman::p_load(janitor, VIM, odds.n.ends, blorr, haven, lmtest, broom, tidyverse, 
               jtools, ggeffects, emmeans, gtsummary, gt, table1, DiagrammeR, mice, table1) 

#/////////////PACKAGE FUNCTIONS USED IN THIS ANALYSIS//////////////////////////////////////

# janitor:        clean_names() - Clean column names
# VIM:                   md.pattern() - Visualize missing data patterns (used once)
# odds.n.ends:               odds.n.ends() - Calculate odds ratios, create ROC and probability plots
# blorr:                         blr_test_hosmer_lemeshow() - Hosmer-Lemeshow goodness-of-fit test
# haven:                            read_xpt() - Import BRFSS SAS XPT files
# lmtest:                               lrtest() - Likelihood ratio tests for model comparison
# broom:                                   tidy(), augment() - Tidy model outputs and create augmented datasets
# tidyverse:                                dplyr, ggplot2, tidyr, stringr, etc. - All data manipulation and visualization
# jtools:                                summ() - Enhanced regression summaries with confidence intervals
# ggeffects:                         ggpredict(), plot() - Create predicted probability plots with confidence intervals
# emmeans:                         emmeans() - Estimated marginal means for post-hoc comparisons
# gtsummary:                    tbl_regression(), tbl_uvregression(), tbl_merge() - Create publication-ready tables
# gt:                       gtsave(), as_gt() - Save tables as HTML files
# table1:             table1() - Create descriptive Table 1 with counts and proportions
# DiagrammeR:     grViz() - Create flowchart diagram
#mice         Missing data plot

# Set theme
theme_set(theme_minimal())
```

# Import dataset and select variables of interest upon entry to avoid unnecessary columns

```{r}
#read 2024 BRFSS data and select potential variables of interest
df1 <- read_xpt('/Users/colemaher/Downloads/LLCP2024.XPT ') %>%
  select("EDUCA", "_SEX", "_IMPRACE", "_AGEG5YR", "FLUSHOT7") %>% 
  # Select these variables
    clean_names()  # Clean column names
        
# Initial data inspection
summary(df1)
md.pattern(df1, rotate.names=TRUE) #missingness isn't a problem. I could evaluate it more broadly, 
#but I am going to leave it be (*complete case analysis*)
```

## Clean up outcome variable (DV = flushot7)

```{r}
# Clean flu shot variable (1 = Yes, 0 = No)
df2 <- df1 %>%
  filter(flushot7 %in% 1:2) %>%  # Keep only valid responses
  mutate(
    flu_shot = case_when(
      flushot7 == 1 ~ 1L,  # Yes
      flushot7 == 2 ~ 0L   # No
    ),
    .keep = "unused"  # Remove original flushot7 column
  ) %>%
  mutate(educa = case_when(educa == 1 ~ 1, #make the first three levels of education (from BRFSS) a single level of non-hs graduation
                          educa == 2 ~ 1, 
                          educa == 3 ~ 1,
                          educa == 4 ~ 2, #make hs grad second level
                          educa == 5 ~ 3, #make some college third level
                          educa == 6 ~ 4, #make college grad (or more) fourth
                          educa > 6 ~ NA)) %>% #otherwise get rid of it
  mutate(educ = factor(educa,
                      levels = 1:4, #create a new ed variable with it factored at these levels
                      labels = c("Some High School or Less", 
                                 "High School Graduate", 
                                 "1-3 years of College or Technical School", 
                                 "College Graduate or more")))

summary(df2) #check missingness
```

## Rename columns to simplify and drop NA values, compare length of data frames

```{r}
# Rename variables and remove missing values
df3 <- df2 %>%
  rename( #rename variables
    education = educa,
    race = imprace,
    age_group = ageg5yr
  ) %>%
  drop_na()  # Remove missing values



## Factor the age variable categories and remove all missingness in age

# Convert categorical variables to factors with proper labels
df3 <- df3 %>%
  mutate(
    age_group = factor(age_group,
                      levels = 1:13,
                      labels = c("18-24", "25-29", "30-34", "35-39", 
                                "40-44", "45-49", "50-54", "55-59", 
                                "60-64", "65-69", "70-74", "75-79", "80+")),
 sex = factor(sex, levels = 1:2, labels = c("Male", "Female")),
    flu_shot = factor(flu_shot, levels = 0:1, labels = c("No", "Yes")),
        race = factor(race, levels = 1:6, labels = c("White, Non-Hispanic", "Black, 
                                                   Non-Hispanic", "Asian, Non-Hispanic", 
                                               "American Indian/Alaskan Native, Non-Hispanic", 	
                                            "Hispanic" , "Other race, Non-Hispanic")) ) %>% 
  #we have missing values = 14 in the age group variable, so these will be removed
  filter(age_group != 14) 

sum(is.na(df3$age_group)) #should be

summary(df3)

# Verify no missing values and check data structure
cat("Final sample size:", nrow(df3), "\n") 
```

# Run first, simplest model

```{r}
# Model with age group only
model1 <- glm(flu_shot ~ age_group, data = df3, family = binomial)

# Enhanced model summary
summ(model1, confint = TRUE, ci.width = 0.95)  # jtools enhanced model summary

# Odds ratios with confidence intervals
print("Model 1")
odds.n.ends(model1, rocPlot = TRUE, predProbPlot = TRUE)
```

## Multicolinearity and Outlier Analysis

```{r}
# Outlier analysis using Cook's Distance
model1_augmented <- augment(model1, type.residuals = "deviance")

# Identify influential observations
influential_obs <- model1_augmented %>%
  mutate(observation = row_number()) %>%
  filter(.cooksd > 4/n())  # Cutoff for cooks

cat("Influential observations (Cook's D > 4/n):", nrow(influential_obs), "\n")

# Plot Cook's Distance
ggplot(model1_augmented, aes(x = seq_along(.cooksd), y = .cooksd)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 4/nrow(model1_augmented), 
             color = "red", linetype = "dashed") +
  labs(x = "Observation", y = "Cook's Distance",
       title = "Cook's Distance Plot") +
  theme_minimal()
```

# Create a multivariate model2 (with race and sex) and test it against our univariate model1

```{r}
# Multivariate model
model2 <- glm(flu_shot ~ age_group + race + sex, 
              data = df3, 
              family = binomial)

# Model summary 
summ(model2, confint = TRUE, ci.width = 0.95)

# Likelihood ratio test comparing models
  lrtest_result <- lrtest(model1, model2)
  print(lrtest_result)

summ(model2, vifs = TRUE)
blr_test_hosmer_lemeshow(model2)

# Odds ratios with confidence intervals
odds.n.ends(model2, rocPlot = TRUE, predProbPlot = TRUE)
```

## Evaluate Potential for Confounding by Race, Sex, and Education

```{r}
# ////////
# ASSESSMENT OF CONFOUNDING: Compare Crude vs Adjusted Models
# //////

# Extract age group ORs from a model (function)
extract_age_ors <- function(model, model_name) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(str_detect(term, "age_group")) %>%
    mutate(
      age_group = gsub("age_group", "", term),
      or = estimate,
      ci_low = conf.low,
      ci_high = conf.high,
      model = model_name
    ) %>%
    select(age_group, or, ci_low, ci_high, model)
}

# Extract ORs from both models
crude_ors <- extract_age_ors(model1, "Crude (Unadjusted)")
adj_ors <- extract_age_ors(model2, "Adjusted (Race + Sex)")

# Combine and calculate percentage change
confounding_assessment <- crude_ors %>%
  full_join(adj_ors, by = "age_group", suffix = c("_crude", "_adj")) %>%
  mutate(
    percent_change = ((or_adj - or_crude) / or_crude) * 100,
    abs_percent_change = abs(percent_change),
    meaningful_confounding = abs_percent_change > 10  # 10% rule
  ) %>%
  select(age_group, or_crude, or_adj, percent_change, meaningful_confounding)

# Print results
cat("\nCONFOUNDING ASSESSMENT: Age Group ORs\n")
cat(rep("=", 60), "\n")
cat("Age Group | Crude OR | Adjusted OR | % Change | >10% Change?\n")
cat(rep("-", 60), "\n")

for(i in 1:nrow(confounding_assessment)) {
  cat(sprintf("%-10s %-10.2f %-13.2f %-10.1f%% %-15s\n",
              confounding_assessment$age_group[i],
              confounding_assessment$or_crude[i],
              confounding_assessment$or_adj[i],
              confounding_assessment$percent_change[i],
              ifelse(confounding_assessment$meaningful_confounding[i], "YES", "no")))
}

# Summary statistics
cat("\nSUMMARY:\n")
cat(rep("=", 40), "\n")
cat(sprintf("Mean absolute percentage change: %.1f%%\n", 
            mean(abs(confounding_assessment$percent_change), na.rm = TRUE)))
cat(sprintf("Maximum percentage change: %.1f%%\n", 
            max(abs(confounding_assessment$percent_change), na.rm = TRUE)))
cat(sprintf("Number of age groups with >10%% change: %d\n", 
            sum(confounding_assessment$meaningful_confounding, na.rm = TRUE)))
```

# Build Models with interaction and without interaction term

```{r}
# ///////////
# WITHOUT interaction
# ///////////
model_main <- glm(flu_shot ~ age_group + race + sex + educ, 
                  family = "binomial", data = df3)

# ///////////
# WITH interaction between age group and education
# ///////////
model_interaction <- glm(flu_shot ~ age_group + educ + race + sex + age_group*educ, 
                         family = "binomial", data = df3)

# Test whether effect modification is present (compare groups with and without interaction term)

lrtest(model_main, model_interaction)

# Print Model Summary for Interaction
summary(model_interaction)
```

# Stratify groups by Education Level and observe odds

```{r}

# ///////////
# Some HS or Less
HS.Less <- glm(flu_shot ~ age_group + race + sex, 
               family = "binomial", data = df3 %>% filter(educ == "Some High School or Less"))
summary(HS.Less)
odds.n.ends(HS.Less)
# ///////////
# High School Graduate
HS.Grad <- glm(flu_shot ~ age_group + race + sex, 
               family = "binomial", data = df3 %>% filter(educ == "High School Graduate"))
summary(HS.Grad)
odds.n.ends(HS.Grad)

# ///////////
# Some College
CL.Some <- glm(flu_shot ~ age_group + race + sex, 
               family = "binomial", data = df3 %>% filter(educ == "1-3 years of College or Technical School")) 
summary(CL.Some)
odds.n.ends(CL.Some)

# ///////////
# College Graduate
CL.Grad <- glm(flu_shot ~ age_group + race + sex, 
               family = "binomial", data = df3 %>% filter(educ == "College Graduate or more"))
summary(CL.Grad)
odds.n.ends(CL.Grad)
```

# Pull estimated marginal means

```{r}
#Run formal Estimated Marginal Means
emm_results <- emmeans(model_interaction, 
                       specs = ~ age_group | educ,  # by age within each education
                       type = "response")  # returns probabilities
print(emm_results)
```

# Create Univariate and Multivariate Tables:

```{r}
# UNIVARIATE (UNADJUSTED) ANALYSIS Table

univ_table <- df3 %>%
  dplyr::select(age_group, flu_shot) %>%
  tbl_uvregression(
    method = glm,
    y = flu_shot,
    method.args = list(family = binomial),
    exponentiate = TRUE,
    label = list(age_group = "Age Group")
  ) %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table 1. Unadjusted Odds Ratios for Influenza Vaccination**") %>%
  bold_labels() %>%
  add_n() %>%
  add_nevent()


# MULTIVARIATE (ADJUSTED) ANALYSIS - MODEL 2 (Age only, but adjusted for race/sex) Table

adj_table_model2 <- tbl_regression(
  model2,
  exponentiate = TRUE,
  label = list(age_group = "Age Group"),
  include = c(age_group),  # Only show age_group, but model includes race/sex
  estimate_fun = function(x) style_number(x, digits = 2),
  pvalue_fun = function(x) style_pvalue(x, digits = 3)
) %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table 2. Adjusted Odds Ratios for Influenza Vaccination**") %>%
  bold_labels() %>%
  add_n() %>%
  add_nevent()
```

#Combine the Tables into a single adjusted and unadjusted table

```{r}
# Create a combined table showing both unadjusted and adjusted ORs
combined_table <- tbl_merge(
  tbls = list(univ_table, adj_table_model2),
  tab_spanner = c("**Unadjusted OR**", "**Adjusted OR**")
) %>%
  modify_caption("**Table 2. Comparison of Unadjusted and Adjusted Odds Ratios for Influenza Vaccination**")

print(combined_table)
```

#Create Stratified Table: 

```{r}
# Clean names
clean_stratified_table <- function(model, model_name) {
  tbl_regression(
    model,
    exponentiate = TRUE,
    label = list(age_group = "Age Group"),
    include = c(age_group),  # Only show age_group
    estimate_fun = function(x) style_number(x, digits = 2),
    pvalue_fun = function(x) style_pvalue(x, digits = 3)
  ) %>%
    modify_header(estimate = paste0("**", model_name, "**")) %>%
    modify_column_hide(columns = c(ci)) %>%
    modify_column_unhide(columns = c(estimate, p.value))
}

# Create tables for each stratified model
strat_tables <- list(
  clean_stratified_table(HS.Less, "OR"),
  clean_stratified_table(HS.Grad, "OR"),
  clean_stratified_table(CL.Some, "OR"),
  clean_stratified_table(CL.Grad, "OR")
)

# Merge stratified tables
stratified_combined <- tbl_merge(
  tbls = strat_tables,
  tab_spanner = c("Some HS or Less", "HS Graduate", "Some College", "College Grad+")
) %>%
  modify_caption("**Table 3. Adjusted Odds Ratios for Influenza Vaccination, Stratified by Education Level**") %>%
  modify_header(label = "**Variable**") %>%
  modify_footnote(estimate = "Odds ratios adjusted for race and sex within each education stratum") %>%
  bold_labels()

print(stratified_combined)

# ///////////
# SAVE TABLES
# ///////////

# Save tables as HTML
gtsave(as_gt(combined_table), "table2_combined.html")
gtsave(as_gt(stratified_combined), "table3_stratified.html")
```

#Figure 1: Flowchart 

```{r}
#build each node , then assign them names given sample size at each df level
grViz(diagram = "digraph flowchart{ # gives beginning of flowchart
      
      node [fontname = Helvetica, shape = rectangle, fontsize = 15] 
    
      node1 [label = '@@1']  # starting number 
      node2 [label = '@@2']  # number after exclusion 1
      node3 [label = '@@3']  # number after exclusion 2
      
      node1 -> node2 -> node3
}
      [1]: 'Initial BRFSS Sample \\n (n = 457,670)'
      [2]: 'Second Sample after excluding missing or invalid flu shot response \\n (n = 413,304) \\n (-44,366)'
      [3]: 'Final Sample after missing or invalid values excluded \\n (n = 405,844) \\n (-7,460)'
      ")
```

#Building confounding plot

```{r}

confounding_plot <- confounding_assessment %>%
  filter(age_group != "18-24") %>%
  ggplot(aes(x = age_group, y = percent_change)) +
  geom_col(aes(fill = meaningful_confounding), alpha = 0.7, width = 0.6) +
  geom_hline(yintercept = c(-10, 10), linetype = "dashed", color = "red", alpha = 0.5) +
  geom_hline(yintercept = 0, color = "black") +
  labs(
    title = "Percentage Change in Age Group ORs After Adjustment",
    subtitle = "Red dashed lines indicate ±10% threshold for meaningful confounding",
    x = "Age Group",
    y = "Percentage Change in OR",
    fill = ">10% Change?"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c(`NO CONFOUNDING` = "steelblue", `CONFOUNDING` = "firebrick"), 
                    drop = FALSE  # This forces both levels to appear in legend
)

print(confounding_plot)

# Save plot
ggsave("confounding_change_plot.png", confounding_plot, width = 10, height = 6)
```

Build interaction plot for the effect modification using predicted probabilities with 95% CI

```{r}
# Create predicted probabilities
predicted_probs <- ggpredict(model_interaction, 
                             terms = c("age_group", "educ"),  
                             # age_group on x-axis, educ as color grouping
                             type = "fixed")
p <- plot(predicted_probs) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 11),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  ) +
  # construct lines and points
  geom_line(linewidth = 1.0) +    
  geom_point(size = 2.5) +      
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), 
              # 95% CI around each line for each group
              alpha = 0.2) +
  labs(
    title = "Predicted Probability of Flu Vaccination",
    subtitle = "Showing age effects stratified by education level",
    x = "Age Group",
    y = "Predicted Probability",
    color = "Education Level",
    fill = "Education Level"
  )

print(p)

# Save plot
ggsave("vaccination_plot_large_elements.pdf", p, width = 12, height = 8)
```

# Table 1

```{r}
#just build the thing
label(df3$age_group) <- "Age Group)"
label(df3$sex) <- "Sex"
label(df3$race) <- "Race"
label(df3$educ) <- "Education Level"

table1( ~ age_group + educ + race + sex | flu_shot, overall = "Total", rowlabelhead = "Variable", df3) 
table1 <- table1( ~ age_group + educ + race + sex | flu_shot, overall = "Total", rowlabelhead = "Variable", df3) 
```
